Profiles of Wind Veer and Wind Speed in Turbulent Ekman Flow

### 2022-03-08 write1/manuscript.tex
some updates in the manuscript (Hauke's comments) and further polishing); first results for ReD1600_50 from hawk.hlrs

\date{\scriptsize Draft - Version \today}
\title{Wind veer and and Speed in Turbulent Ekman Flow}
\author{Cedrick Ansorge, Hauke Wurps}

@@

Wind veer away from its geostrophic direction is
(i) due to direct frictional effects in the very vicinity of the surface and
(ii) due to turbulence which exerts indirect frictional effects; these effects
cause a slow-down of the mean wind reducing the Coriolis force thus turning the wind
in favor of the pressure gradient force. 
%%
Not only does the veering set the frame of reference for surface layer theory, it also has
effects at small and large scales \todo{Insert: Brown, 2005; Svensson and Holtslag, 2009}.
%
From a large-scale perspective, the veering of wind across the planetary boundary layer determines
the amount of cross-isobaric mass-flux, commonly referred to as 'Ekman pumping' \citep{ekman:AMA1905},
and it is thus a key factor in the life-cycle of large-scale synoptic systems.
%
% %% This shear is probably not related to Ekman-turning 
% On the mesoscale, shear in the surface layer a crucial factor for the initiation and sustainment of strong convective events (REFERENCE?).
% 
Within the atmospheric boundary layer (ABL), directional shear of the wind in
the upper part of the surface layer may cause a systmatic yaw for tall wind power generation devices
where blades reach into the Ekman layer, i.e. that part of the boundary layer where the wind starts to turn;
an exact estimate of such effects is critical in the site assessments for wind farms
\todo{References: Calaf et al., 2010 Phys Fluids; Mirocha et al., 2018 Wind Energy Science} . 
%
\par
%
In the planetary boundary layer, wind veer is characterized by the surface veering angle $\alpha$
defined as the angle between the negative surface shear stress $\tau_\SFC$ and the geostrophic wind.
%
Surface veering $\alpha$ and geostrophic drag $Z\equiv u_\star/G$, where $u_\star\equiv\sqrt{\nu|\tau_\SFC|}$,
uniquely determine the surface drag $\tau_\SFC$ in a turbulent Ekman flow. 
%
In any quantitative description of the surface layer, the friction velocity $u_\star$ is the dynamic scale
and $\alpha$ defines the alignment of the frame of reference.
%
Knowledge about $u_\star$ and $\alpha$ is thus a pre-requisite for any quantitative theory of the surface layer
and a \citet{rossby:PPO1935} constrained the two parameters based on integral relations in the ABL.
%
Asymptotic similarity theory was later used by \citet{tennekes:JAS1973, blackadar:JAS1968},
and--based on his seminal direct numerical simulations (DNS) of Ekman flow--,~\cite{spalart:JFM1989} suggested
a modification to take into account effects of low to intermediate Reynolds numbers.
Later on, constants were re-evaluated with a focus on the atmospheric boundary layer based on
observations \citep{hogstrom:BM1988,hogstrom:BM1996}
and numerical modelling \citep{spalart:PF2008, spalart:PF2009, ansorge:BM2014, ansorge:BM2019}.
\todo{Refer to Pena et al, 2010} 
%
\par
%
Attempts were also undertaken to obtain profiles of the wind speed: 
One approach is to match the inner and outer layer at a reference height;
\citet{etling:2002, emeis:2018}  (Sec. 21.10; Eq. 21.48) choose the Prandtl-layer height to match
the wind speed profiles, which, however, requires to externally prescribe the veering $\alpha$ at that height. 
A one-dimensional profile with constant veering is given by \citet[Sec. 3; Eq. 3.1-3.19]{emeis:m2007}.

@@

\section{Problem Form
  ulation and Numerical Approach}

@@

@@

a broader scale separation that is made available by the two new simulations for 

@@

turbulent and there is no distinct inner-oouter scale separation.

@@

Also in the outer layer of the flow, $u_\star$ (not the magnitude the
geostrophic wind speed $G$) is the governing velocity scale for inviscid, i.e. $\RE$-independent,

@@

which has indeed already been found by \cite{spalart:JFM1989}, who estimate the constant from

@@

    the turbulent stress $S_\mathrm{turb}^+$ (solid), 

@@

    and the total stress, $S^+=S_\mathrm{visc}^++S_\mathrm{turb}^+$ (dashed)

@@

exhibits universal scaling when considered as $S_\mathrm{visc}^+(z^+)$ (Fig.~\ref{fig:stresses}c); this normalization is also appropriate in the outer layer where the viscous stress

@@

at higher $z^+$ (up to $z^+\approx 50$ for $\RE=1600$; Fig.~\ref{fig:stresses}b).

@@

This can be formulated in terms of the modulus
of the surface friction normalized by the geostrophic wind forcing, $u_\star$ (Eq.~(\ref{eqn:ustar_def}))
and, the direction of surface shear stress, $\alpha_\star$ (Eq.~(\ref{eqn:alpha_def})),
also termed wind veering. 
A non-zero veering of the wind is a rather special case in comparison with most turbulent flows conisdered in an

@@

We compare our DNS data against a semi-empirical drag-law based on integral consideration \cite{spalart:JFM1989}
and find, as demonstrated in previous work \cite{ansorge:BM2014}, excellent agreement in the range

@@

  u_\star= 4\log(\mathbf{Re}_D))-8

@@

While this is not the case here for the general form of the profile, we

@@

In immediate vicinity to the surface, local turbulent mixing cannot occur for the no-slip/no-penetration boundary condition,
and the mean velocity is described by a viscous profile of the form
\begin{equation}
  U^{\alpha_\star+}  = z^+
\end{equation}
where the direction of the velocity points into the exact opposite direction of the wall shear stress $\tau$. 
In absence of roughness elements and for small roughness ($z_0^+<5$), this linear regime is known as
viscous sub-layer \cite{foken:m2002,foken:BM1978}.
%
In fact, this law of the wall has no degree of freedom given the drag, i.e. once $u_\star$ and $\alpha_\star$ are defined. 
%
However, theoretical foundation is lacking for the exact shape of the velocity profile in the buffer layer; though crucial for turbulence
production, it is commonly understood as a transition region between the linear profile at the surface and the logarithmic profile
aloft.
A pure blending from the linear velocity profile into the logarithmic one is, however, not reasonable as both the linear and logarithmic
profile overestimate the velocity in the buffer layer. We therefore introduce a higher-order compensating term into
the linear profile before blending it into the logarithmic law:
\begin{equation}
  U_\text{inner}^{\alpha_\star}^+ = f_{U,\mathrm{visc}}(z^+) = \frac{ z^+ + \gamma_4 (z^+)^{4} + \gamma_6 (z^+)^{6} } { 1 + \gamma_6 /u_{ref} (z^+ )^6}, 
\end{equation} 
where the fourth- and sixth-order terms in the nominator are designed to fit the profile below $y^+\approx 10$ and
the sixth-order correction in the denominator limits the growth of $u_\text{inner}$ for large $y^+$.
%
These correction terms are small for $z^+<5$ where the linear law of the wall
(shown in the inset in Fig.~\ref{fig:u_profiles}) yields an excellent fit of the data; we use here
\begin{equation*}
    \gamma_4 = -3.825 \times 10^{-4};\quad
    \gamma_6 &=&  6.320 \times 10^{-6};\quad 
    u_\mathrm{ref} &=& 1.278 \times 10^{1}.
\end{equation*}

@@

@@

approaches a $\RE$-independent constant around $C_{V0}=Z_\star \sin\alpha -6.1$ at the surface; deviations from

@@

\subsubsection{Discussion}

>>>>

\date{\scriptsize Draft - Version \today}
\title{Profiles of Wind Veer and Wind Speed in Turbulent Ekman Flow}
\author{Cedrick Ansorge, Hauke Wurps}

@@

Wind veer away from its geostrophic direction is
(i) due to direct frictional effects in the very vicinity of the surface and
(ii) due to turbulence which exerts indirect frictional effects; these effects
cause a slow-down of the mean wind reducing the Coriolis force thus turning the wind
in favor of the pressure gradient force. 
%%
Not only does the veering set the frame of reference for surface layer theory, it also has
effects at small and large scales from large-scale dispersion via plume spreading to
cyclone spin-down \citep{svensson:BM2009}
and on the capabilities of data assimilation and accuracy of surface flux estimates \citep{brown:QJR2005}. 
%
From a large-scale perspective, the veering of wind across the planetary boundary layer determines
the amount of cross-isobaric mass-flux, commonly referred to as 'Ekman pumping' \citep{ekman:AMA1905},
and it is thus a key factor in the life-cycle of large-scale synoptic systems.
%
% %% This shear is probably not related to Ekman-turning 
% On the mesoscale, shear in the surface layer a crucial factor for the initiation and sustainment of strong convective events (REFERENCE?).
% 
Within the atmospheric boundary layer (ABL), directional shear of the wind in
the upper part of the surface layer may cause a systmatic yaw for tall wind power generation devices
where blades reach into the Ekman layer, i.e. that part of the boundary layer where the wind starts to turn;
an exact estimate of such effects is critical in the site assessments for wind farms
\citep{calaf:PF2010, mirocha:WES2018}. 
%
\par
%
In the planetary boundary layer, wind veer is characterized by the surface veering angle $\alpha$
defined as the angle between the negative surface shear stress $\tau_\SFC$ and the geostrophic wind.
%
Surface veering $\alpha$ and geostrophic drag $Z\equiv u_\star/G$, where $u_\star\equiv\sqrt{\nu|\tau_\SFC|}$,
uniquely determine the surface drag $\tau_\SFC$ in a turbulent Ekman flow. 
%
In any quantitative description of the surface layer, the friction velocity $u_\star$ is the dynamic scale
and $\alpha$ defines the alignment of the frame of reference.
%
Knowledge about $u_\star$ and $\alpha$ is thus a pre-requisite for any quantitative theory of the surface layer, 
and \citet{rossby:PPO1935} constrained the two parameters based on integral relations in the ABL.
%
Asymptotic similarity theory was later used by \citet{tennekes:JAS1973, blackadar:JAS1968},
and--based on his seminal direct numerical simulations (DNS) of Ekman flow--,~\cite{spalart:JFM1989} suggested
a modification to take into account effects of low to intermediate Reynolds numbers.
Later on, constants were re-evaluated with a focus on the atmospheric boundary layer based on
observations \citep{hogstrom:BM1988,hogstrom:BM1996}
and numerical modelling \citep{spalart:PF2008, spalart:PF2009, ansorge:BM2014, ansorge:BM2019}.
\todo{Refer to Pena et al, 2010} 
%
\par
%
Attempts were also undertaken to obtain profiles of the wind speed: 
One approach is to match the inner and outer layer at a reference height;
\citet{etling:2002, emeis:2018}  (Sec. 21.10; Eq. 21.48) choose the Prandtl-layer height to match
the wind speed profiles, which, however, requires to externally prescribe the veering $\alpha$ at that height. 
A one-dimensional profile with constant veering is given by \citet[Sec. 3; Eq. 3.1-3.19]{emeis:m2007}.

@@

\section{Problem Formulation and Numerical Approach}

@@

Another common measure of scale separation is the Reynolds number defined by the laminar Ekman layer
thickness $D=\sqrt{2\nu/f}$, 
  \RE_D = \frac{G D} {\nu } = \sqrt{2 \RE_\Lambda}. 

@@

a broader scale separation that is made available by the two new simulations at increased Reynolds number  

@@

turbulent and there is no distinct inner-outer scale separation.

@@

Also in the outer layer of the flow, $u_\star$ (and not the magnitude of the
geostrophic wind $G$) governs the inviscid, i.e. $\RE$-independent,

@@

which has indeed already been found by \cite{spalart:JFM1989}, who estimates the constant from

@@

    the turbulent stress $S_\mathrm{turb}^+$ (dashed), 

@@

    and the total stress, $S^+=S_\mathrm{visc}^++S_\mathrm{turb}^+$ (solid)

@@

exhibits universal scaling when considered as $S_\mathrm{visc}^+(z^+)$ (Fig.~\ref{fig:stresses}a); this normalization is also appropriate in the outer layer where the viscous stress

@@

at higher $z^+$ for higher Reynolds number (up to $z^+\approx 50$ for $\RE=1600$; Fig.~\ref{fig:stresses}c).

@@

This can be formulated by the normalized surface friction, $u_\star$ (Eq.~(\ref{eqn:ustar_def}), also termed geostrophic drag), 
and the direction of surface shear stress, $\alpha_\star$ (Eq.~(\ref{eqn:alpha_def}), also termed wind veer). 
A non-zero veering of the wind is a rather special case in comparison with most turbulent flows considered in an

@@

We compare our DNS data against a semi-empirical drag-law based on integral consideration \citep{spalart:JFM1989}
and find, as demonstrated in previous work \citep{ansorge:BM2014}, excellent agreement in the range

@@

  Z_\star= 4\log(\mathbf{Re}_D))-8

@@

While this is not the case here for the general form of the profile, the

@@

In immediate vicinity to the surface, local turbulent mixing cannot occur for the no-slip/no-penetration boundary condition,
and the mean velocity is described by a viscous profile of the form
  U^{\alpha_\star+}  = z^+
where the direction of the velocity points into the exact opposite direction of the wall shear stress $\tau$. 
In absence of roughness elements and for small roughness ($z_0^+<5$), this linear regime is known as
viscous sub-layer \cite{foken:m2002,foken:BM1978}.
In fact, this law of the wall has no degree of freedom given the drag, i.e. once $u_\star$ and $\alpha_\star$ are defined. 
However, theoretical foundation is lacking for the exact shape of the velocity profile in the buffer layer; though crucial for turbulence
production, it is commonly understood as a transition region between the linear profile at the surface and the logarithmic profile
aloft.
A pure blending from the linear velocity profile into the logarithmic one is, however, not reasonable as both the linear and logarithmic
profile overestimate the velocity in the buffer layer.
We therefore introduce a two-step correction procedure, accounting for the smaller-than linear growth beyond $y^+\approx 5$, and assuring smooth matching with the logarithmic law at $y^+=40$:
  U_\text{inner}^{\alpha_\star}^+ = \frac{z^+}{1+c_1 (y^+)^2} +  (c_2 y^+ - a_\mathrm{match} )\frac{1+tanh\left[0.2(y^+-22)\right]}{2} + c_3 e^{-c_4(y^+-22)^2}. \label{eqn:u_inner} 
We use here
\begin{equation*}
  c_1 &=& 0.00185;\quad
  c_2 &=& 0.195;\quad 
  c_3 &=& 0.4 ; \quad
  c_4 &=& 0.35. 
\end{equation*}
The second and third terms on the right hand side vanish for $y^+\ll 22$, and $c_1=0.00185$
implies an approximately 5\% correction at $y^+=5$ and an 18.5\% correction at $y^+=10$.
The second and third term on the R.H.S. of eq.~(\ref{eqn:u_inner}) are an empirical fit to
the velocity profiles observed in the buffer layer and appear independent of the Reynolds number
for the range observed here.
The coefficient $a_\mathrm{match}$, which has no effect in the viscous sublayer, is then used to
match this formulation to the logarithmic law employed above. 

@@

For this logarithmic law, $a_\mathrm{match}=3.569861$ for a matching at $y^+=40$. 

@@

approaches a $\RE$-independent constant around $C_{V0}=Z_\star \sin\alpha=6.1$ at the surface; deviations from

@@

\section{Discussion}
\subsection{} 
